\bf Distortion corrected kernel density estimate on Riemannian manifolds

### 2023-06-05 jcgs/03-chap3-kderm.Rmd
Formatting and citations

When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$. Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh. One could easily apply KDE to the output vectors $\pmb{y}$. However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is 
$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,$$
where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$. By a change of variables, this would imply the following density for $\pmb{y}$,
\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
\end{equation}
where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$. It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant. Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g. ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.  Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].

@@

In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample. Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case. For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$. We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
\begin{equation}
\begin{aligned}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1]; 
(v) \sup K(\|\pmb{y}\|)=K(0).
\end{aligned}
\end{equation}

@@

For data $\pmb{y}_i\in\mathbb{R}^d$ with $i=1,\dots,N$ and assuming a bandwidth matrix $r\pmb{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
\end{equation}

The intuition behind this estimator is very clear for a uniform kernel. The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein. Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.

<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density. To make this clear consider the simpler case of computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term. A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density of the manifold itself.  Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density on its own. This motivates a kernel density estimate that corrects for the distortion induced by $\psi$.-->

@@

For kernel density estimation on a known manifold, @Pelletier2005-vu propose the following estimator,
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
\end{equation}

where $d_g(\pmb{p}, \pmb{p}_i)$ denotes the geodesic distance between two points on the manifold $\pmb{p}$ and $\pmb{p}_i$ and $\theta_{\pmb{p}_i}(\pmb{p})$ is known as the volume density function. The intuition behind the term $K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)$ is relatively clear. For example, for a uniform kernel, the estimator at point $\pmb{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$. However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance. An additional technical assumption is that $r$ is less than the injectivity radius of the manifold. A definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix. For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\pmb{p}$ is so large that some points `fall inside' the ball more than once. For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere. This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.

The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$. However, the volume of this ball must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center. The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.

More formally, the volume density function can be explained as follows. Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space,  $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$. Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and traveling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time. Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system. The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system. In proving that the estimator integrates to 1, the volume density function cancels with the term $|\pmb{H}(\pmb{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\pmb{p}_i$. For more on the volume density function, see @Gallot2004-rc and @Le_Brigant2019-lj.

@@

To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric. Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$. <!--The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities. For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems. This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions. To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point. These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm. <!--For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$. -->

While full details are provided in @Perrault-Joncas2013-pq, we briefly describe the Learn Metric algorithm here. There are four main steps in the algorithm. First, a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold. Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph. Third, a dimension reduction method is applied to obtain the output embedding $\pmb{y}_1,\dots,\pmb{y}_n$. Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator). Full details on these four steps are provided in \autoref{alg:learnmetric}. This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*. Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.

As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension $d$. In this case, the ranks of the matrices $\pmb{H}_i$ are equal to $d$. Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension $d$ globally into $d$-dimensional Euclidean space. In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into $d$-dimesional Euclidean space. In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used (See Chapter 3 of @Lee2007-wq for more about intrinsic dimension estimation).  The *R* library *intRinsic* [@Denti2021-qc] implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm EMBED }
  \OptParameter{ manifold learning parameters in EMBED }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;

  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
  $$
  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
  $$
  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;

  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;

>>>>

When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\bm{p}$ to $d$-dimensional vectors $\bm{y}$. Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh. One could easily apply KDE to the output vectors $\bm{y}$. However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
$$Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,$$
where $f$ is the probability density function on the manifold, $\bm{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$. By a change of variables, this would imply the following density for $\bm{y}$,
\begin{equation}
\label{eq:changevar}
Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
\end{equation}
where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\bm{H}(\bm{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$. It can be seen that the density in terms of $\bm{y}$, differs from the density on the manifold due to the term $|\bm{H}(\bm{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant. Since $\psi$ will typically be a non-linear function, the value of $|\bm{H}(\bm{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\bm{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of $|\bm{H}(\bm{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g. ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\bm{y}$ to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space $\bm{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.  Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].

@@

In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\bm{p}$ in general, with $\bm{p}_1,\dots,\bm{p}_n$ denoting the observed sample. Often $\bm{p}_i$ will be high-dimensional vectors such that $\bm{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case. For instance, $\bm{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each $\bm{p}_i$ only require some sense of distance between the 'input' points, $d(\bm{p}_i,\bm{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\bm{y}_1,\dots,\bm{y}_n$, where $\bm{y}_i\in\mathbb{R}^d$. We will denote this embedding as $\bm{y}_i=\psi(\bm{p}_i)$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
\begin{equation}
\begin{aligned}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=1;
(ii) \int_{\mathbb{R}^{d}} \bm{y} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\bm{y}\|^{2} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\bm{y}\|)=K(0).
\end{aligned}
\end{equation}

@@

For data $\bm{y}_i\in\mathbb{R}^d$ with $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\bm{y}$ is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\bm{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\bm{y}-\bm{y}_i\|}{r}\right).
\end{equation}

The intuition behind this estimator is very clear for a uniform kernel. The density at a point $\bm{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\bm{y}$, times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius $r$ centered at $\bm{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein. Kernel density estimators of this form can and have been applied directly on the output embedding $\bm{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.

<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density. To make this clear consider the simpler case of computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term. A similar notion applies to a manifold embedding so that the density of the output vectors $\bm{y}_i$ differs from the density of the manifold itself.  Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density on its own. This motivates a kernel density estimate that corrects for the distortion induced by $\psi$.-->

@@

For kernel density estimation on a known manifold, @Pelletier2005-vu propose the following estimator,
\begin{equation}
\label{eq:denriem}
\hat{f}(\bm{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\bm{p}_i}(\bm{p})} K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right),
\end{equation}

where $d_g(\bm{p}, \bm{p}_i)$ denotes the geodesic distance between two points on the manifold $\bm{p}$ and $\bm{p}_i$ and $\theta_{\bm{p}_i}(\bm{p})$ is known as the volume density function. The intuition behind the term $K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)$ is relatively clear. For example, for a uniform kernel, the estimator at point $\bm{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\bm{p}$. However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance. An additional technical assumption is that $r$ is less than the injectivity radius of the manifold. A definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix. For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\bm{p}$ is so large that some points `fall inside' the ball more than once. For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere. This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.

The inclusion of the volume density function  $\theta_{\bm{p}_i}(\bm{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\bm{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\bm{p}$. However, the volume of this ball must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center. The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.

More formally, the volume density function can be explained as follows. Consider the exponential map around $\bm{p}$, given by $exp_{\bm{p}}(\bm{q})$, mapping vectors in the tangent space,  $\bm{v}\in T_{\bm{p}}M$, to points on the manifold, $\bm{q}\in M$. Loosely, $\bm{v}$ `points' in the direction of the geodesic between $\bm{p}$ and $\bm{q}$ and traveling along this geodesic at uniform speed $\|\bm{v}\|$ takes place in one unit of time. Now, consider a chart $\varphi$ mapping points in the neighborhood of $\bm{p}$, via the inverse of the exponential map, to these $\bm{v}$ vectors, expressed in some local coordinate system. The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system. In proving that the estimator integrates to 1, the volume density function cancels with the term $|\bm{H}(\bm{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\bm{p}_i$. For more on the volume density function, see @Gallot2004-rc and @Le_Brigant2019-lj.

@@

To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\bm{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric. Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space $T_{\bm{p}} M$ for every point $\bm{p} \in M$. <!--The inner product between two tangent vectors $u,v \in T_{\bm{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities. For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems. This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\bm{y}_1,\bm{y}_2\dots,\bm{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\bm{y}$, angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions. To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\bm{y}_1,\bm{y}_2\dots,\bm{y}_n$ with $d\times d$ positive definite matrices, $\bm{H}_1,\bm{H}_2\dots,\bm{H}_n$, at each data point. These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm. <!--For example, the angle between $\bm{p}_j$ and $\bm{p}_k$ at $\bm{p}_i$ depends (up to a first order approximation) on the inner product $(\bm{y}_j-\bm{y}_i)'\bm{H}^{-1}_i(\bm{y}_k-\bm{y}_i)$ rather than the usual Euclidean inner product $(\bm{y}_j-\bm{y}_i)'(\bm{y}_k-\bm{y}_i)$. -->

While full details are provided in @Perrault-Joncas2013-pq, we briefly describe the Learn Metric algorithm here. There are four main steps in the algorithm. First, a weighted neighborhood graph is constructed, with edges between $\bm{p}_i$ and $\bm{p}_j$ when $\bm{p}_i$ is a K-nearest neighbor of $\bm{p}_j$ or vice versa, and edge weights depending on the distance between $\bm{p}_i$ and $\bm{p}_j$ on the manifold. Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph. Third, a dimension reduction method is applied to obtain the output embedding $\bm{y}_1,\dots,\bm{y}_n$. Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator). Full details on these four steps are provided in \autoref{alg:learnmetric}. This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*. Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.

As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension $d$. In this case, the ranks of the matrices $\bm{H}_i$ are equal to $d$. Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension $d$ globally into $d$-dimensional Euclidean space. In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into $d$-dimesional Euclidean space. In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used (See Chapter 3 of @Lee2007-wq for more about intrinsic dimension estimation).  The *R* library *intRinsic* [@Denti2021-qc] implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\bm{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\bm{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\bm{H}_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm EMBED }
  \OptParameter{ manifold learning parameters in EMBED }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\bm{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\bm{x}_i-\bm{x}_j\|^2)$ for data points $\bm{x}_i,\bm{x}_j \in \mathbb{R}^s$;

  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
  $$
  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
  $$
  where $\widetilde{D}=diag{\widetilde{W}\bm{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\bm{1}}$;

  \STATE Embed all data point $\bm{X}\in \mathbb{R}^s$ to embedding coordinates $\bm{Y}=(\bm{y}^1,\dots,\bm{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;

  \STATE Obtain the $d\times d\times N$ tensor $\bm{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\bm{Y}$ with elements in $\bm{\tilde{H}}$ found by
  $$
    \bm{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\bm{y}^l \odot \bm{y}^k\right) - \bm{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \bm{y}^k\right) - \bm{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \bm{y}^l\right)\right],
  $$
  where $l,k=1,\dots,d$, $\bm{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;

  \STATE For each embedding data point $\bm{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\bm{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\bm{H}^{\cdot\cdot i}}$ below
  $$
    \bm{H}_i = \bm{U} \bm{\Lambda}_d^{-1} \bm{U}^\prime,
  $$
  where $[\bm{U}, \bm{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\bm{H}^{\cdot\cdot i}}$, $\bm{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\bm{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.
